---
title: "Effect modification and interaction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect modification and interaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# The estimates table names each row by the effect measure, the contrast it
# compares, and the subgroup it belongs to, which needs more than the console's
# default 80 columns to print without wrapping.
options(width = 100)
```

Two questions get asked with the same words and answered with different
machinery. This vignette separates them, shows a dataset on which they give
different answers, and explains why that is the correct behavior rather than a
discrepancy to be reconciled.

```{r setup}
library(propensity)

# The weight functions report which exposure they found and how they read it.
# That is useful once and repetitive here, so quiet it for the vignette.
options(propensity.quiet = TRUE)
```

## Two different questions

**Effect modification** asks whether the effect of a treatment `A` differs
across strata of some variable `V`. It is a question about *conditioning*. You
split the population by `V`, estimate the effect of `A` in each part, and
compare. `V` is never intervened on and needs no causal role at all. It can be
a biomarker, a region, a calendar year, or anything else you can group people
by. `ipw()` answers this with `.by`.

**Interaction** asks what happens under a joint intervention on two treatments
`A` and `E`. It is a question about *setting both*. Both treatments have to be
identifiable: each needs positivity and a sufficient adjustment set, and the
second needs a model conditioning on the first. `propensity` answers this with
`joint_exposure()` for a declared crossing, or with `joint_wt_models()` and
`wt_joint()` for a pair of treatment models.

The distinction is not a technicality about which function to call. The two
questions have different answers, and a variable can modify an effect while
being useless as a target of intervention.

There is a genuine coincidence result, but it is about one variable analyzed
both ways. Conditioning on `V` and intervening on `V` answer the same question
when `V` is independent of whatever carries the effect heterogeneity, or when
there is no heterogeneity to carry. That says nothing about how effect
modification by one variable relates to interaction with a different one. Those
two are free to take any pair of values, as the sections below show.

## A dataset where they diverge

The design is a surrogate modifier. A covariate `l` carries genuine
heterogeneity in the effect of `a`. A second variable `v` is strongly
associated with `l` but does nothing at all: it does not appear in the outcome,
and setting it would change nothing. The second treatment `e` depends on `l`
and on `a`.

```{r simulate}
set.seed(2024)
n <- 2000

l <- rbinom(n, 1, 0.5)
v <- factor(
  ifelse(rbinom(n, 1, 0.15 + 0.7 * l) == 1, "high", "low"),
  levels = c("low", "high")
)
a <- rbinom(n, 1, plogis(-0.4 + 0.9 * l))
e <- rbinom(n, 1, plogis(-0.5 + 1.0 * l + 0.8 * a))
y <- 1 + 0.4 * a + 0.3 * e + 0.5 * l + 1.2 * a * l + 0.5 * a * e + rnorm(n)

dat <- data.frame(l, v, a, e, y)
```

Two facts about this data generating process matter later. The effect of `a` is
larger when `l` is 1, because of the `1.2 * a * l` term. And `v` is a marker for
`l` and nothing more:

```{r composition}
round(prop.table(table(dat$v, dat$l), 1), 3)
```

Each stratum of `v` is about 84% one value of `l`. That composition is the whole
of what `v` contributes.

By construction the additive interaction between `a` and `e` is 0.50, the
coefficient on `a * e`, since the outcome is linear and `l` averages out. The
effect modification by `v` works out to about 0.91: the effect of `a` is 0.75
where `l` is 0 and 2.04 where `l` is 1, and the two strata of `v` mix those in
opposite proportions. Two different numbers, both true, describing two different
things.

## Effect modification with `.by`

Weights handle the confounding of `a` by `l`, and the outcome model is the
marginal structural model in `a` and `v`. `.by` then reports the effect within
each stratum along with the difference between them.

```{r em}
ps_a <- glm(a ~ l, data = dat, family = binomial())
w_a <- wt_ate(ps_a)

em_mod <- lm(y ~ a * v, data = dat, weights = w_a)
ipw(ps_a, em_mod, .by = v)
```

The effect of `a` is about 0.92 where `v` is low and about 1.95 where `v` is
high, and the difference between the strata is about 1.03. Nothing here is
wrong. The effect of `a` really is larger among people with `v = high`, and if
you wanted to know where the treatment does the most good, this is the answer.

What it does not say is that `v` causes anything. Every bit of the difference
comes from the strata having different mixes of `l`. Give someone `v = high`
without changing their `l` and their response to `a` is unchanged. This is why
`.by` requires no assumptions about `V`: it makes no claims about `V` either.

Stabilizing the weights on the modifier is worth doing here, and `?ipw`
describes the recipe under **Effect modification**.

## Interaction with a joint intervention

Now the other question. Both treatments are intervened on, so both need
treatment models. The second conditions on the first, which is the factorization
a joint weight actually has. `wt_joint()` reads each component's exposure type
off the component, so nothing has to say which of the two is which.

```{r joint}
ps_e <- glm(e ~ a * l, data = dat, family = binomial())

models <- joint_wt_models(a = ps_a, e = ps_e)
w_joint <- wt_joint(w_a, wt_ate(ps_e))

joint_mod <- lm(y ~ a * e, data = dat, weights = w_joint)
joint_fit <- ipw(models, joint_mod)
joint_fit
```

The surface reports the four counterfactual cell means, each treatment's simple
effects within the levels of the other, and one interaction row. The interaction
is about 0.50, which is the truth the data was built with.

A declared crossing gives the same rows through one model over the cells rather
than two models of the treatments:

```{r joint-exposure, eval = FALSE}
joint <- causalgenerics::joint_exposure(a = dat$a, e = dat$e)
ps_cells <- nnet::multinom(joint ~ l, trace = FALSE)
w_cells <- wt_ate(predict(ps_cells, type = "probs"), joint)
ipw(ps_cells, lm(y ~ a * e, data = dat, weights = w_cells))
```

Prefer the two-model route when the treatments call for different adjustment
sets, or when the dependence of the second on the first is what you want to
model. Prefer the declared crossing when one model over the cells is the natural
specification.

## Why the two numbers differ

The effect modification contrast is about 1.03. The interaction is about 0.50.
These are not two estimates of one quantity that ought to agree.

The first compares the effect of `a` between two groups of people you did not
create, whose difference is their composition. The second compares the effect of
`a` between two worlds you did create, one in which everyone gets `e` and one in
which nobody does. The first is a difference of two conditional effects, each
computed in its own covariate distribution. The second is a double difference
within a single population.

Neither number constrains the other, and this data generating process shows that
in both directions if you adjust one piece of it at a time.

Make `v` independent of `l` and change nothing else. The effect modification
contrast goes to zero, because the strata of `v` now have the same mix of `l`
and there is nothing left for stratifying on `v` to find. The interaction stays
at 0.50, because `l` had nothing to do with it. Someone who read the effect
modification result as the interaction would conclude there was no interaction
when there is one of 0.50, which is why this is the case where conflating the
two misleads most, not the case where they agree. At n = 2000 a single run
scatters around that zero with a standard error of about 0.11, so read it as an
expectation rather than as a number one sample will land on.

Now put `v` back and drop the `a * e` term instead, keeping `a * l`. The reverse
happens: the interaction goes to zero while the effect modification contrast
stays near 0.95. The effect of `a` still differs across strata of `v`, and
setting `e` still does nothing to it.

So the pair can be about (0, 0.50), or about (0.95, 0), or, as in the data
above, (1.03, 0.50). Asking which one is "the" interaction is asking the wrong
question.

The practical consequence is about what each result licenses. A large effect
modification contrast tells you where to look for people who benefit. Only the
interaction tells you what happens if you act on the second thing.

## Interaction is symmetric

The interaction is reported once, under the first treatment's framing. That is
not a choice about which treatment matters more; the two framings are the same
number. The change in `a`'s effect when `e` goes from 0 to 1 equals the change in
`e`'s effect when `a` goes from 0 to 1:

```{r symmetry}
est <- joint_fit$estimates
est[est$effect == "diff", c("contrast", "group", "estimate")]

pick <- function(contrast, group) {
  est$estimate[
    est$effect == "diff" & est$contrast == contrast & est$group == group
  ]
}

c(
  # a's effect at e = 1 minus a's effect at e = 0
  a_framing = pick("a: 1 vs 0", "e = 1") - pick("a: 1 vs 0", "e = 0"),
  # e's effect at a = 1 minus e's effect at a = 0
  e_framing = pick("e: 1 vs 0", "a = 1") - pick("e: 1 vs 0", "a = 0"),
  reported = pick("a: 1 vs 0", "e = 1 vs e = 0")
)
```

The three agree exactly rather than approximately, because all of them are
contrasts over the same block of cell means in one stacked system. Reporting
both framings would be one quantity under two labels, so only the first
treatment's is shown. The equality is pinned in the package's tests.

## Odds ratios and effect modification

With a binary outcome, `.by` reports the risk difference and the log risk ratio
for the strata and their contrast, but the log odds ratio only for the overall
rows:

```{r noncollapsible}
dat$yb <- rbinom(
  n,
  1,
  plogis(-1.2 + 0.5 * a + 0.3 * e + 0.6 * l + 0.9 * a * l)
)

emb_mod <- glm(
  yb ~ a * v,
  data = dat,
  family = quasibinomial(),
  weights = w_a
)
ipw(ps_a, emb_mod, .by = v)
```

This is deliberate. The odds ratio is noncollapsible: a stratum-specific odds
ratio can differ from the population odds ratio, and from the odds ratio in
another stratum, purely because the outcome is distributed differently in each
stratum. Two strata with identical risk differences and identical risk ratios can
still have different odds ratios. A difference of two odds ratios therefore moves
with the outcome's baseline risk whether or not the effect is modified at all,
and reporting one under the heading of effect modification would invite exactly
the reading it does not support. The same reasoning removes the odds ratio from
the joint intervention surface entirely.

If you want an odds ratio, take the overall one. If you want effect
modification, take it on a collapsible scale.

## Guardrails on joint weights

A joint weight is not the product of any two weights. The weight for a joint
intervention on `A` and `E` is `1 / [f(A | L) f(E | A, L)]`, and the second
factor conditions on the first treatment. The product of two *marginal* weights,
`1 / [f(A | L) f(E | L)]`, is a different quantity. `joint_wt_models()` refuses
it:

```{r guardrail, error = TRUE}
marginal_e <- glm(e ~ l, data = dat, family = binomial())
joint_wt_models(a = ps_a, e = marginal_e)
```

The refusal exists because nothing downstream can catch the mistake. Both
products are ordinary vectors of positive numbers, both fit an outcome model
without complaint, and both return estimates with standard errors that look
entirely reasonable.

How much damage it does depends on the data, and in a way that is worth knowing.
When the confounder enters the outcome additively, the marginal product is very
nearly unbiased for the interaction: the errors in the four cell means cancel in
the double difference. The failure appears once the confounder also modifies the
treatment effects, which is the situation an interaction analysis is usually run
in. In simulations for this guardrail, with a confounder modifying both effects
and an odds ratio of about 4.5 for the first treatment's effect on the second, a
true interaction of 0.75 came back as 0.39, an attenuation of roughly half.
Under stronger dependence and stronger confounding the estimate changes sign
altogether, returning a negative interaction where the truth is positive. The
magnitude scales with the outcome, so no single number characterizes it; the
direction of the failure is the part to remember.

Two further requirements follow from the same factorization. The dependence of
the second treatment on the first should be modeled flexibly rather than as a
single additive term. An additive term satisfies the check, since the check reads
the formula, but a badly modeled dependence biases the joint weight just as
invisibly as omitting the term. Above, `e ~ a * l` lets the dependence differ by
`l` rather than forcing one shift. And a continuous component must be stabilized:
the unstabilized density ratio has a heavy right tail on its own, and a product
inherits it.

The check lives on the container. `wt_joint()` multiplies whatever it is given,
so building weights by hand and fitting an outcome model directly bypasses the
guardrail. Route joint analyses through `joint_wt_models()`.

## Two notes on labels

Grouped rows are labeled with treatment levels, as in `"a: 1 vs 0"` and
`"e = 1"`. Those labels come from the data as it was coded, so they are not
portable across two codings of one intervention: the same analysis with `a` as a
factor of `"no"` and `"yes"` produces different label text for identical
estimates. This matters when pooling, since pooled results are keyed by their
labels. Pooling across multiply imputed datasets is safe whenever the analyses
share one data preparation, which they normally do.

On the joint dose route, where the second treatment is continuous, the
coefficient surface accepts any coding, and the coding shows up in the
coefficient names the rows carry. Each row reports the fit's own coefficient
under whatever coding it was given, so this is not the previous paragraph's
situation of one estimate under two labels: an ordered factor contributes a
polynomial contrast column, and the row named for it reports that coefficient,
which is a different quantity from the indicator's. The vocabulary surface,
whose rows claim to be the effect at a dose of zero and the slope at the
reference level, refuses a non-indicator coding outright rather than making
those claims of a column that does not support them.

## Learning more

- `?ipw`, sections **Effect modification** and **Joint exposures**, for the
  reported row sets and their restrictions
- `?wt_joint` and `?joint_wt_models` for the factorization and its checks
- `?wt_ate`, section **Stabilization**, for stabilizing on a modifier
